Verified Nanopasses for Compiling Conditionals

### 2025-06-19 paper.tex
lift locals

\lstset{%
language=Python,
basicstyle=\ttfamily\small,
morekeywords={match,case,bool,int,let,begin,if,else,closure},
deletekeywords={},
escapechar=|,
mathescape=true,
columns=flexible,
%moredelim=[is][\color{red}]{~}{~},
showstringspaces=false,
%inputencoding=utf8
literate={→}{{\to}}1,
}

@@

In this section we describe how we formalize the syntax and semantics
of the source language and four intermediate languages in Agda:
\begin{enumerate}
\item Integers, Booleans, \texttt{let}, and \texttt{if}
\item Monadic Normal Form
\item Imperative with Structured Control Flow
\item A C-like language with \emph{goto}
\item X86 assembly but with variables
\end{enumerate}
These languages and there semantics are quite standard. However, there
are many other intermediate languages in the literature to choose
from, so why these languages, and why in this particular order?  The
answer is that they happen to represent the images of the four
nanopasses.  So then the question becomes, why do we choose to order
the nanopasses in this way? The main thing to consider is that some
passes make later passes easier because they remove a language feature
that would have been complicated to handle. For example, the Lift
Locals pass removed \texttt{let}, and I found \texttt{let} difficult
to deal with in the proof of correctness for Explicate Control
(\texttt{let} is relatively easy to handle in the code for Explicate
Control). The other thing to consider is that some compiler passes use
certain language features in their generated code, so the passes that
remove those language features better come later in the ordering.

@@

\subsection{The source language}

The source language is a language with integers, Booleans,
\texttt{Let} binding, and conditional expressions. We choose a couple
representative primitive operators: \texttt{Sub} for integer
subtraction and \texttt{Eq} for equality on integers and
Booleans.\footnote{Adding more primitive operators is straightforward
but uninteresting.}  The \texttt{Read} operation requests an integer
input from the user of the program.

@@

\subsection{Monadic normal form}

The first intermediate language restricts the arguments of the
primitives \lstinline{Sub} and \lstinline{Eq} to be \emph{atomic}
expressions, which we define to include variables and literals. This
restriction gets us closer to x86, where instruction arguments may
only be immediates (literals) or registers.

@@

\subsection{Structured imperative IR}

@@

@@

>>>>

\lstset{%
language=Python,
basicstyle=\ttfamily\small,
keywordstyle=\ttfamily\small,
%morekeywords={match,case,bool,int,let,begin,if,else,closure},
%deletekeywords={locals},
%escapechar=|,
mathescape=true,
columns=flexible,
%moredelim=[is][\color{red}]{~}{~},
showstringspaces=false,
%inputencoding=utf8
literate={→}{{\to}}1,
}

@@

In this section we describe how we formalize the syntax and semantics
of the source language and four intermediate languages in Agda:
\begin{enumerate}
\item Integers, Booleans, \texttt{let}, and \texttt{if}
\item Monadic Normal Form
\item Imperative with Structured Control Flow
\item A C-like language with \emph{goto}
\item X86 assembly but with variables
\end{enumerate}
These languages and there semantics are quite standard. However, there
are many other intermediate languages in the literature to choose
from, so why these languages, and why in this particular order?  The
answer is that they happen to represent the images of the four
nanopasses.  So then the question becomes, why do we choose to order
the nanopasses in this way? The main thing to consider is that some
passes make later passes easier because they remove a language feature
that would have been complicated to handle. For example, the Lift
Locals pass removes \texttt{let}, and I found \texttt{let} difficult
to deal with in the proof of correctness for Explicate Control
(\texttt{let} is relatively easy to handle in the code for Explicate
Control). The other thing to consider is that some compiler passes use
certain language features in their generated code, so the passes that
remove those language features better come later in the ordering.  For
example, the Closure Conversion pass (not discussed here) generates
code with \texttt{let}, so Closure Conversion comes before Lift
Locals.

@@

\subsection{The Source Language}

The source language includes integers, Booleans, \texttt{Let} binding,
and conditional expressions. We choose a couple representative
primitive operators: \texttt{Sub} for integer subtraction and
\texttt{Eq} for equality on integers and Booleans.\footnote{Adding
more primitive operators is straightforward but uninteresting.}  The
\texttt{Read} operation requests an integer input from the user of the
program.\footnote{\texttt{Read} is necessary to prevent students from
interpreting the programs instead of compiling them.}

@@

\subsection{Monadic Normal Form}

The first intermediate language restricts the arguments of the
primitives \lstinline{Sub} and \lstinline{Eq} to be \emph{atomic}
expressions, which we define to include variables and literals. This
restriction gets us closer to x86, where instruction arguments may
only be immediates (literals), registers, or memory accesses.

@@

\subsection{Structured Imperative IR}

@@

We implement the Remove Complex Operands pass with the following
\lstinline{rco} function in Agda. This pass is responsible for making
sure that the arguments of primitive operations, such as
\lstinline{Sub} and \lstinline{Eq}, are atomic expression (variables
or literals).

\begin{lstlisting}
rco : Exp $→$ Mon
\end{lstlisting}

Most of the cases of \lstinline{rco} are trivial, but let us focus on
the case for subtraction. We shall start with a strawman
translation. We recursively compile the subexpressions $e_1$ and
$e_2$. The results might be complex (not atomic), so we
\lstinline{let} binding them both and replace them in the subtraction
with variables (de Bruijn indices 0 and 1). Note that we must shift
the variables in $e_2$ up by one because we inserted a new
\lstinline{let} above $e_2$. (In general, \lstinline{shift-mon m c}
increases by one each variable in $m$ that is greater or equal to the
cutoff $c$.)

\begin{lstlisting}
rco (Sub e$₁$ e$₂$) =
  Let (rco e$₁$)
   (Let (shift-mon (rco e$₂$) 0)
    (Sub (Var 1) (Var 0)))
\end{lstlisting}

However, the above translation inserts more \lstinline{let} bindings
than are strictly necessary. If $e_1$ and/or $e_2$ compile to an atom,
then there is no need to bind them to a \lstinline{let}.  The
following is an alternative translation for \lstinline{Sub} that
includes four different cases, depending on whether \lstinline{rco e$₁$}
and \lstinline{rco e$₂$} are atomic.

\begin{lstlisting}
rco (Sub e$₁$ e$₂$)
    with rco e$₁$ | rco e$₂$
... | m$₁$ | m$₂$
    with atomic? m$₁$ | atomic? m$₂$
... | yes (atomic a$₁$) | yes (atomic a$₂$) =
      Sub a$₁$ a$₂$
... | no cmplx$₁$ | yes (atomic a$₂$) =
      Let (rco e$₁$) (Sub (Var zero) (shift-atm a$₂$ 0))
... | yes (atomic a$₁$) | no cmplx$₂$ =
      Let (rco e$₂$) (Sub (shift-atm a$₁$ 0) (Var 0))
... | no cmplx$₁$ | no cmplx$₂$ = 
      Let m$₁$
        (Let (shift-mon m$₂$ 0)
          (Sub (Var 1) (Var 0)))
\end{lstlisting}

(It would be nice to collapse the four cases into one, but I have not
yet figured out how to do that while accounting for the necessary
shifting of de Bruijn indices.)

The correctness theorem for Remove Complex Operands says that for any
program $e$, interpreting $\mathtt{rco}\,e$ with input $s$ yields the
same output as directly interpreting $e$ with the same input.

\begin{lstlisting}
rco-correct : $∀$ (e : Exp) (s : Inputs)
  $→$ interp-LMonIf (rco e) s $≡$ interp-LIf e s 
\end{lstlisting}

The main lemma is that \lstinline{rco} followed by
\lstinline{interp-mon} is equivalent to \lstinline{interp-exp}.

\begin{lstlisting}
rco-correct-exp : ∀ (e : Exp) (ρ : Env Value)
  → interp-mon (rco e) ρ ≡ interp-exp e ρ
\end{lstlisting}

The proof is by structural induction on $e$, but it requires the
following lemmas regarding the shifting of de Bruijn indices.
The idea is that the result of looking up variable $x$ in
environment ($ρ₁$ ++ $ρ₂$) produces the same result as
shifting the variable by one with the cutoff set to the
the length of $ρ₁$ and looking it up in the environment
($ρ₁$ ++ v $∷$ $ρ₂$), for any value $v$.

\begin{lstlisting}
nth-++-shift-var : ($ρ₁$ $ρ₂$ : List A)(v : A)(x : Id)
  $→$ nth ($ρ₁$ ++ v $∷$ $ρ₂$) (shift-var x (length $ρ₁$))
    $≡$ nth ($ρ₁$ ++ $ρ₂$) x

interp-shift-atm : (a : Atm)(v : Value)($ρ₁$ $ρ₂$ : Env Value)
  $→$ interp-atm (shift-atm a (length $ρ₁$)) ($ρ₁$ ++ v $∷$ $ρ₂$) 
    $≡$ interp-atm a ($ρ₁$ ++ $ρ₂$)

interp-shift-mon : (m : Mon)(v : Value)($ρ₁$ $ρ₂$ : Env Value)
  $→$ interp-mon (shift-mon m (length $ρ₁$)) ($ρ₁$ ++ (v $∷$ $ρ₂$))
    $≡$ interp-mon m ($ρ₁$ ++ $ρ₂$)
\end{lstlisting}

@@

The Lift Locals pass is responsible for replacing all of the
\lstinline{let} expressions with assignment statements.  After this
pass, variables are still represented by de Bruijn indices but they
are all bound by the \lstinline{Program} AST node.

The \lstinline{lift-locals} function delegates to the recursive
function \lstinline{lift-locals-mon}, where all the work is done.
That function returns a number $n$ that counts the number of
\lstinline{let} bindings in the expression, and the translated
expression.

\begin{lstlisting}
lift-locals : Mon → Imp-Prog
lift-locals m
    with lift-locals-mon m
... | n , e = Program n e    
\end{lstlisting}

\noindent The case of \lstinline{lift-locals-mon} for \lstinline{Let}
is the most interesting. One can imagine that the result of
\lstinline{lift-locals-mon m$₁$} is a sequence of $i$ imaginary
\lstinline{let}'s surrounding \lstinline{e$₁$}. Similarly for 
\lstinline{lift-locals-mon m$₁$}.
The translation result for this \lstinline{Let} should be
a sequence of $i + j + 1$ imaginary
\lstinline{let}'s, followed by an assignment
of \lstinline{e$₁$} to the appropriate variable, and then the
code for \lstinline{e$₂$}. However, we need to shift
the variables in \lstinline{e$₁$} up by $j + 1$ because
we are moving all of the $j$ \lstinline{let}'s for
\lstinline{e$₂$} above \lstinline{e$₁$}, as well as the
current \lstinline{let}. Thus, we need another function
for shifting de Bruijn indices by an arbitrary amount.

\begin{lstlisting}
shifts-imp-exp : Imp-Exp $→$ $ℕ$ $→$ $ℕ$ $→$ Imp-Exp
\end{lstlisting}

The variables in \lstinline{e$₂$} also need to be shifted.  They are
shifted up by $i$ because we are moving all of the $i$
\lstinline{let}'s for \lstinline{e$₁$} above
\lstinline{e$₂$}. Regarding the variable that serves as the left-hand
side of the assignment, its de Bruijn index is $i + j$ because the
original index for this \lstinline{Let} was $0$ but we have moved $i +
j$ other \lstinline{let}s above this variable occurence.

\begin{lstlisting}
lift-locals-mon (Let m$₁$ m$₂$)
    with lift-locals-mon m$₁$
... | i , e$₁$
    with lift-locals-mon m$₂$
... | j , e$₂$ =
    let e$′₁$ = shifts-imp-exp e$₁$ 0 (suc j) in
    let e$′₂$ = shifts-imp-exp e$₂$ j i in
    suc (i + j) , Assign (i + j) e$′₁$ e$′₂$
\end{lstlisting}

The case of \lstinline{lift-locals-mon} for \lstinline{If} is also
complicated because there are three subexpressions that must all be
shifted appropriately.

The correctness theorem for Lift Locals says that if a program output
is $v$ given input $s$, then Lift Locals produces a program that
outputs the same value given the same input.

\begin{lstlisting}
lift-locals-correct : $∀$(m : Mon)(s : Inputs)(v : Value)
  $→$ interp-LMonIf m s $≡$ just v
  $→$ interp-imp (lift-locals m) s v
\end{lstlisting}

This theorem is a corollary of the following main lemma, which is
proved by structural induction on the expression \lstinline{m}.
Considerable care is needed in the statement of this lemma so that the
induction hypotheses can be used. Recall that when interpreting an
expression, the environment $ρ$ provides the meaning for the
\emph{free variables} in the expression.  As the interpreter descends
under \lstinline{let} bindings, it pushes values on the front of the
environment. However, after Lift Locals, all of the variables are
bound at the top, so the initial environment won't be empty, but
instead will have entries for all of the variables (we arbitrarily
initialize them to $0$). So in general, we can separate the
environment into two halves, $ρ₁$ and $ρ₂$, where $ρ₂$ is for the free
variables in \lstinline{m} whereas \lstinline{$ρ₁$ ++ $ρ₂$} is for the
free variables in the result expressoin \lstinline{e} of
\lstinline{lift-locals-mon m}. The length of $ρ₁$ corresponds to the
number returned by \lstinline{lift-locals-mon m}. Also, note that
because \lstinline{Mon} is a pure language, \lstinline{e} contains a
single assignment to each variable, so $ρ₂$ is unchanged by the
execution of \lstinline{e}. ($ρ₁$ is changed as each of its variables
are assigned to for the first time.)

\begin{lstlisting}
lift-locals-mon-correct : $∀$(m : Mon)(e : Imp-Exp)
    (s $s′$ : Inputs)(v : Value) ($ρ₁$ $ρ₂$ : Env Value)
  $→$ interp-mon m $ρ₂$ s $≡$ just (v , $s′$)
  $→$ lift-locals-mon m $≡$ (length $ρ₁$ , e)
  $→$ $∃$[ $ρ′₁$ ] (s , $ρ₁$ ++ $ρ₂$) $⊢$ e $⇓$ v $⊣$ ($s′$ , $ρ′₁$ ++ $ρ₂$)
      × length $ρ₁$ $≡$ length $ρ′₁$
\end{lstlisting}

The main lemma relies on more lemmas about shifting de Bruijn indices
that generalize our previous ones to arbitrary amounts of shifting.

\begin{lstlisting}
$⇓$shifts : $∀${e}{v}{s $s′$}{$ρ₁$ $ρ′₁$ $ρ₂$ $ρ₃$ $ρ′₃$} 
  $→$ (s , $ρ₁$++$ρ₃$) $⊢$ e $⇓$ v $⊣$ ($s′$ , $ρ′₁$++$ρ′₃$)
  $→$ length $ρ′₁$ $≡$ length $ρ₁$
  $→$ (s , $ρ₁$++$ρ₂$++$ρ₃$) $⊢$
       shifts-imp-exp e (length $ρ₁$) (length $ρ₂$)
       $⇓$ v $⊣$ ($s′$ , $ρ′₁$++$ρ₂$++$ρ′₃$)
\end{lstlisting}
